CT simulation

### 2021-05-12 analysis/paper/garvey_ct_simulation.Rmd
start thinking about simulations

>>>>

Ben's explorations of simulating artefacts over multiple generations. THis looks at one attribute only, so far

```{r}
library(tidyverse)

data <- read.csv(here::here("analysis/data/garvey_data.csv"))

# how many generations?
generations <- 100

# how many artefacts per generation?
n_artefacts_per_generation <- 30

# what CVs to simulate with?
cvs <- seq(1, 50, by = 5) / 100
length(cvs) # 50 different CV values

# which dimension of the artefact?
dimen <- names(data)[1]

# create an empty list object to store the results
generations_by_cv_lst <- vector("list", length = length(cvs))
# start a new list to hold all the generations for this cv value
generations_lst <- vector("list", length = generations)

for(cv_i in 1:length(cvs)){

  # assign the CV value for generating the simulated data
  SD <- cvs[cv_i]

# for each generation, simulate data based on the previous generation
for(g in 1:generations){

  # for generation 1, get the first generation from the archaeological data

  if(g == 1){ 
    d        <- na.omit(data[, dimen])    # select the data set to sample
    generation_seed <-  sample(d, 1, replace = TRUE) 
    generations_lst[[g]] <-  
     abs( rnorm(n_artefacts_per_generation, 
             mean(generation_seed), 
             SD * mean(generation_seed)))

  } else {

  # for generations 2, 3, 4, etc. simulate the data from the previous generation
  previous_generation_index <-  g - 1
  previous_generation_data <-  generations_lst[[previous_generation_index]]

  # and generate a new set of random normal variables from the previous generation
  generations_lst[[g]] <- 
    abs(rnorm(n_artefacts_per_generation, 
            mean(previous_generation_data), 
            SD * mean(previous_generation_data)))
  }
}

# make into a single vector, and store in list element
generations_by_cv_lst[[cv_i]] <- (generations_lst)

# take a look 
#  generations_by_cv_lst

# plot some
library(viridis)

plot_list <- 
  map(1:length(cvs),
      function(cv){
library(ggridges)
tibble(`001` =    generations_by_cv_lst[[cv]][[1]],
       `025` =    generations_by_cv_lst[[cv]][[25]],
       `050` =    generations_by_cv_lst[[cv]][[50]],
       `100` =    generations_by_cv_lst[[cv]][[100]]) %>% 
  pivot_longer(everything()) %>% 
  mutate(name = fct_rev(name)) %>% 
ggplot()  +
  geom_density_ridges(aes(x = value, 
                          y = name,
                          fill = name)) +
  theme_minimal()+
  scale_fill_viridis_d() +
  guides(fill = "none") +
  labs(x = paste0("CV = ", cvs[cv]))
})

library(cowplot)
plot_grid(plotlist = plot_list)

# plot all generations

map(1:length(cvs),
    function(cv){
  map(1:generations, 
    function(g){
      generations_by_cv_lst[[cv]][[g]]
    }) %>% 
  bind_cols() %>% 
  pivot_longer(everything()) %>% 
  mutate(gen = parse_number(str_remove_all (name, "\\.*"))) %>% 
  arrange(gen) %>% 
  group_by(gen) %>% 
  summarise(mean_val = mean(value))
    }) %>% 
  bind_rows(.id = "CV") %>% 
  mutate(CV = as.factor(cvs[as.numeric(CV)])) %>% 
  ggplot() +
  aes(gen, 
      mean_val,
      colour = CV,
      group = CV) +
  geom_line() +
  scale_color_viridis_d()
```

Ben exploring ABC for parameter estimation, can we use this instead of distance metrics to find a good fit with our data?

```{r}
#-------------------------------------------------------------------
# from https://theoreticalecology.wordpress.com/2012/12/02/the-easyabc-package-for-approximate-bayesian-computation-in-r/

library(EasyABC)

# assuming the data ...
data =  d

# we want to use ABC to infer the parameters that were used.
# we sample from the same model and use mean and variance
# as summary statstitics for the model and the data.

# observed summary statistics
summarydata = c(mean(data), sd(data))

# stochastic model generates a sample for given par and returns summary statistics
model <- function(par){ 
  samples <- rnorm(100, mean =par[1], sd = par[2])
  return(c(mean(samples), sd(samples)))
}

# prior list - this is where we assert our opinions about the data-generating
# process
prior_list <- list(c("unif", 0, 10),
                   c("unif", 0, 10))

# call to EasyABC with the ABC-MCMC algorithm Marjoram, P.; Molitor, J.; Plagnol, V. & 
# Tavare, S. (2003) Markov chain Monte Carlo without likelihoods. Proc. Natl. Acad. Sci. USA, 100, 15324-15328.
# with some automatic adjustment options 
ABC_Marjoram_original <- ABC_mcmc(method = "Marjoram", 
                                model = model, 
                                prior = prior_list, 
                                summary_stat_target = summarydata, 
                                n_rec = 10000)

str(ABC_Marjoram_original)
par(mfrow=c(2,1))
hist(ABC_Marjoram_original$param[5000:10000,1], 
     main = "Posterior for mean")
hist(ABC_Marjoram_original$param[5000:10000,2], 
     main = "Posterior for standard deviation")

##------------------------------------------------------
library(EasyABC)

# assuming the data are 10 samples of a normal distribution
# with mean 5.3 and sd 2.7
data =  rnorm(10, mean =5.3, sd = 2.7)

# we want to use ABC to infer the parameters that were used.
# we sample from the same model and use mean and variance
# as summary statstitics for the model and the data.

# observed summary statistics
summarydata = c(mean(data), sd(data))

# stochastic model generates a sample for given par and returns summary statistics
model <- function(par){ 
  samples <- rnorm(10, mean =par[1], sd = par[2])
  return(c(mean(samples), sd(samples)))
}

# call to EasyABC with the ABC-MCMC algorithm Marjoram, P.; Molitor, J.; Plagnol, V. & 
# Tavare, S. (2003) Markov chain Monte Carlo without likelihoods. Proc. Natl. Acad. Sci. USA, 100, 15324-15328.
# with some automatic adjustment options 
ABC_Marjoram_original <- 
  ABC_mcmc(method="Marjoram", 
           model=model, 
           prior=list(c("unif",0,10),
                      c("unif",1,5)), 
           summary_stat_target=summarydata, 
           n_rec = 10000)

str(ABC_Marjoram_original)
par(mfrow=c(2,1))
hist(ABC_Marjoram_original$param[5000:10000,1], main = "Posterior for mean")
hist(ABC_Marjoram_original$param[5000:10000,2], main = "Posterior for standard deviation")

#-------------------------------------------------------------------
#

library(abc)
data(human)

par(mfcol = c(1,3), mar=c(5,3,4,.5))
boxplot(stat.3pops.sim[,"pi"]~models, main="Mean nucleotide diversity")
boxplot(stat.3pops.sim[,"TajD.m"]~models, main="Mean Tajima's D")
boxplot(stat.3pops.sim[,"TajD.v"]~models, main="Var in Tajima's D")

cv.modsel <- cv4postpr(models, 
                       stat.3pops.sim, 
                       nval=5, tol=.01, 
                       method="mnlogistic")
s <- summary(cv.modsel)

modsel.ha <- postpr(target = stat.voight["hausa",], # observed data summary stats
                    index = models,                 # model ID, to go with sumstat
                    sumstat =  stat.3pops.sim,      # summary stats for simulated data 
                    tol=.05, 
                    method="mnlogistic")

modsel.it <- postpr(stat.voight["italian",], models, stat.3pops.sim, tol=.05, method="mnlogistic")
modsel.ch <- postpr(stat.voight["chinese",], models, stat.3pops.sim, tol=.05, method="mnlogistic")
summary(modsel.ha)

# ## an artifical example
ss <- cbind(runif(1000),
            rt(1000,df=20))
# ss is a df of 1000 rows and two columns
pp <- 
postpr(target=c(3), 
       index=c(rep("a",500),rep("b",500)),
       sumstat=ss[,1], tol=.1, method="rejection")

summary(pp)

```
